Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2020-02-04 manuscript/climate_sensitivity_draft.Rmd
working on text

One fundamental question regarding forest responses to drought is what drives the observed tendency for large trees to suffer more during drought. @bennett_larger_2015 was the first to show greater growth reductions for larger trees on a global scale, and numerous subsequent studies have reinforced this finding (*e.g.*, @stovall_tree_2019; @hacket-pain_consistent_2016], although there are exceptions [@zuleta_drought-induced_2017]. However, it has yet to be resolved which of several potential underlying mechanisms drive this pattern. First, taller trees have a greater biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018], especially during drought [@zhang_size-dependent_2009]. To meet this challenge, hydraulic traits vary in relation to tree height within indivdiuals {**REFS**}, within species {**REFS**}, and across species [@liu_hydraulic_2019]. Despite the tendency for hydrualic traits to become more conservative with height on an individual tree {**REFS**}, the need for greater hydraulic efficiency (*i.e.,* larger xylem conduits) in tall trees [@liu_hydraulic_2019]{**REFS**} may leave them more vulnerable during drought {**REFS-- Mark Olsen--mentioned by Lawren**}. Second, larger trees may have lower drought resistance because of their higher crown positions, where they are exposed to higher solar radiation, greater wind speeds, and lower humidity [@koike_leaf_2001; @kunert_revised_2017 **REFS-KAT**). Subcanopy trees tend to fare better specifically due to the benefits of a buffered environment [@pretzsch_drought_2018]. Potentially counteracting the biophysical challenges faced by large trees, their larger root systems confer a potential advantage in terms of allowing greater access to water; however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure. A final mechanism that could mediate tree size-related responses to drought is how species, and their associated hydraulic traits, are distributed with respect to size [@meakem_role_2018; @liu_hydraulic_2019]. Within a forest community, greater drought vulnerability of larger trees could occur if the larger size classes were dominated by species less adapted to handle drought. Alternatively, larger size classes may be dominated by species that are inherently better adapted to greater biophysical challenges--as is the case in tropical moist forests of Panama, where larger size classes contain greater proportions of deciduous species [@condit_quantifying_2000; @meakem_role_2018]. Understanding the mechanisms driving the greater growth reductions of larger trees during drought will require sorting out the interactive effects of height, canopy position, soil water availability, and species' traits.  {**ask Lawren to check revisions to this paragraph**}

@@

Community-level tree growth responses to all three droughts were modest, with mean resistance values of 0.86, 0.84, and 0.86 for 1966, 1977, and 1999 droughts, respectively (Fig. 1b). In each drought, roughly 30% of the cored trees suffered $\ge$ 30% growth reductions ($Rt \le 0.7$): 29% in 1966, 32% in 1977, and 27% in 1999. However, some individuals exhibited increased growth: ($Rt > 1.0$): 26% in 1966, 22% in 1977, and 26% in 1999.

@@

Resistance was negatively correlated with $ln[TWI]$ (Tables 4-5), negating the hypothesis that trees in moist microsites would suffer less during drought. Nevertheless, we tested for a negative $ln[H] *ln[TWI]$ interaction, which could indicate that smaller trees (with smaller rooting volume) are more susceptible to drought in drier microenvironments that have a deeper water table. This hypothesis was rejected; the $ln[H] *ln[TWI]$ interaction was never signficant and had a consistently positive coefficient (Table 4).

@@

Tree size, microenvironment, **see comment 38** and hydraulic traits shaped tree growth responses across three droughts in a temperate deciduous forest (Table 1). The greater susceptibility of larger trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height [@liu_effect_1993; @stovall_tree_2019]. There was a marginal additional effect of crown exposure, with the highest drought impacts felt by the most exposed and the most suppressed trees. This is consistent with observations of both the former (e.g., [@suarez_factors_2004]; [@scharnweber_confessions_2019]) and the latter (**REFS**). **see comment 39** There was no evidence that soil water availability increased drought resistance; in contrast, trees in wetter topographic positions experienced more growth reduction (consistent with [@zuleta_drought-induced_2017; @stovall_tree_2019]), and larger rooting volum provided no advantage in the drier microenvironments. **see comment 40** The lower drought resistance of larger trees was not driven by a trend of more drought-sensitive species dominating the canopy. Drought sensitivity was also not consistently linked to species' $LMA$, $WD$, or xylem architecture. However, drought resistance was negatively correlated with the leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$) in the top overall model and the top models for two of the three individual droughts [@scoffoni_leaf_2014;@bartlett_correlations_2016; @medeiros_extensive_2019]. This is a novel finding in that $PLA_{dry}$ and $\pi_{tlp}$ have not previously been linked to drought growth responses. The direction of responses was mostly consistent across droughts, supporting the conclusion that they were driven by fundamental physiological mechanisms. However, the strengths of each predictor varied across droughts, indicating that specific drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected. These findings significantly advance our knowledge of the factors that confer vulnerability or resistance on trees during drought.

The droughts considered here were of similar severity (Fig. 1b) and fairly moderate **see comment 41**; droughts of this magnitude have occurred with an average frequency of approximately one per 10-15 years (Fig. 1a, @helcoski_growing_2019). Therefore, we excpect that most species are adapted, and individual trees acclimatized, to survive droughts of this nature when not coupled with other disturbances. **see comment 42,43** While the majority of trees experienced reduced growth, a substantial portion had increased growth (Fig. 1b), **see comment 44** supporting the observation that these droughts did not induce extreme stress on the entire forest. It is likely for this reason, along with other non-climate factors influencing tree growth, that our best models characterize only a modest amount of variation: 11-13% for all droughts combined, and 21-26% for each individual drought (Table 5). The moderate nature of these droughts is helpful for our analysis because it considers only trees that survived all of these droughts, and we lack information on the trees that were killed prior to 2008. These are likely to be relatively modest in number, and local forest monitoring data stretching back to the late 1980s confirms that the 1999 drought did not trigger major declines in tree abundance or biomass (*Anderson-Teixeira et al., in review*).  Thus, the droughts considered here are substantially weaker than others that have triggered massive tree die-off in other regions (e.g., [@allen_global_2010]), many of which have shaped our understanding about the role of tree size [@bennett_larger_2015; @stovall_tree_2019] and-to some extent-traits underlying drought response mechanisms (@greenwood_tree_2017; @anderegg_meta-analysis_2016). This is important because while our results are consistent with findings from more extreme, multiannual droughts, these extremes have not occurred in the Eastern United States similar to the West, and therefore we do not know exactly how forests would respond under higher mortality conditions [@clark_impacts_2016; @druckenbrod_redefining_2019; @berdanier_multiyear_2016]. Given this consistency with our results, however, we believe this is an indication of a more intensive response.

>>>>

One fundamental question regarding forest responses to drought is what drives the observed tendency for large trees to be more affected by drought. @bennett_larger_2015 was the first to show greater growth reductions for larger trees on a global scale, and numerous subsequent studies have reinforced this finding (*e.g.*, @stovall_tree_2019; @hacket-pain_consistent_2016], although there are exceptions [@zuleta_drought-induced_2017]. However, it has yet to be resolved which of several potential underlying mechanisms drive this pattern. First, taller trees have a greater biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018], especially during drought [@zhang_size-dependent_2009]. To meet this challenge, hydraulic traits vary in relation to tree height within indivdiuals {**REFS**}, within species {**REFS**}, and across species [@liu_hydraulic_2019]. Despite the tendency for hydrualic traits to become more conservative with height on an individual tree {**REFS**}, the need for greater hydraulic efficiency (*i.e.,* larger xylem conduits) in tall trees [@liu_hydraulic_2019]{**REFS**} may leave them more vulnerable during drought {**REFS-- Mark Olsen--mentioned by Lawren**}. Second, larger trees may have lower drought resistance because of their higher crown positions, where they are exposed to higher solar radiation, greater wind speeds, and lower humidity [@koike_leaf_2001; @kunert_revised_2017 **REFS-KAT**). Subcanopy trees tend to fare better specifically due to the benefits of a buffered environment [@pretzsch_drought_2018]. Potentially counteracting the biophysical challenges faced by large trees, their larger root systems confer a potential advantage in terms of allowing greater access to water; however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure. A final mechanism that could mediate tree size-related responses to drought is how species, and their associated hydraulic traits, are distributed with respect to size [@meakem_role_2018; @liu_hydraulic_2019]. Within a forest community, greater drought vulnerability of larger trees could occur if the larger size classes were dominated by species less adapted to handle drought. Alternatively, larger size classes may be dominated by species that are inherently better adapted to greater biophysical challenges--as is the case in tropical moist forests of Panama, where larger size classes contain greater proportions of deciduous species [@condit_quantifying_2000; @meakem_role_2018]. Understanding the mechanisms driving the greater growth reductions of larger trees during drought will require sorting out the interactive effects of height, canopy position, soil water availability, and species' traits.  {**ask Lawren to check revisions to this paragraph**}

@@

Community-level tree growth responses to all three droughts were modest, with mean resistance values of 0.86, 0.84, and 0.86 for 1966, 1977, and 1999 droughts, respectively (Fig. 1b). In each drought, roughly 30% of the cored trees experienced $\ge$ 30% growth reductions ($Rt \le 0.7$): 29% in 1966, 32% in 1977, and 27% in 1999. However, some individuals exhibited increased growth: ($Rt > 1.0$): 26% in 1966, 22% in 1977, and 26% in 1999.

@@

Resistance was negatively correlated with $ln[TWI]$ (Tables 4-5), negating the hypothesis that trees in moist microsites would be less affected by drought. Nevertheless, we tested for a negative $ln[H] *ln[TWI]$ interaction, which could indicate that smaller trees (with smaller rooting volume) are more susceptible to drought in drier microenvironments that have a deeper water table. This hypothesis was rejected; the $ln[H] *ln[TWI]$ interaction was never signficant and had a consistently positive coefficient (Table 4).

@@

Tree size, microenvironment, and hydraulic traits shaped tree growth responses across three droughts in a temperate deciduous forest (Table 1). The greater susceptibility of larger trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height [@liu_effect_1993; @stovall_tree_2019]. There was a marginal additional effect of crown exposure, with lowest $Rt$ among the most exposed (dominant) and suppressed trees. There was no evidence that soil water availability increased drought resistance; in contrast, trees in wetter topographic positions experienced more growth reduction (consistent with [@zuleta_drought-induced_2017; @stovall_tree_2019]), and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments. Drought sensitivity was not consistently linked to species' $LMA$, $WD$, or xylem architecture, but was negatively correlated with the leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$) in the top overall model and the top models for two of the three individual droughts [@scoffoni_leaf_2014;@bartlett_correlations_2016; @medeiros_extensive_2019]. This is a novel finding in that $PLA_{dry}$ and $\pi_{tlp}$ have not previously been linked to drought growth responses. The direction of responses was mostly consistent across droughts, supporting the conclusion that they were driven by fundamental physiological mechanisms. However, the strengths of each predictor varied across droughts (Tables 4-5), indicating that specific drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected. These findings significantly advance our knowledge of the factors that confer vulnerability or resistance on trees during drought.

The droughts considered here were of a magnitude that has occurred with an average frequency of approximately one per 10-15 years (Fig. 1a, @helcoski_growing_2019) and had modest impacts on tree growth (Fig. 1b). While the majority of trees experienced reduced growth, a substantial portion had increased growth (Fig. 1b), potentially due to decreased leaf area of competitors during the drought. It is likely because of the moderate severity of these droughts, along with other factors influencing tree growth, that our best models characterize only a modest amount of variation: 11-13% for all droughts combined, and 21-26% for each individual drought (Table 5). While each of these droughts has been linked to tree mortality in the eastern United States [@druckenbrod_redefining_2019], they were substantially weaker than many of the extreme, multiannual droughts that have triggered massive tree die-off in other regions (e.g., [@allen_global_2010]). Such extremes have not occurred in the Eastern United States within the past several decades [@clark_impacts_2016; @druckenbrod_redefining_2019; @berdanier_multiyear_2016], and thus our findings--while generally consistent with findings from more severe droughts [@bennett_larger_2015; @zuleta_drought-induced_2017; @stovall_tree_2019]--apply primarily to moderate-severity droughts.
